**Measuring Depression and Anxiety with 4 items? Adaptation of the PHQ-4 to increase its Sensitivity to Subclinical Variability**

### 2025-06-05 paper/5_Collabra/manuscript.qmd
update
update analysis

  The PHQ-4 is an ultra-brief (4 items) screening questionnaire for depression and anxiety. In this brief report, we test the benefits of adding one additional response option ("Once or twice", in between "Not at all" and "Several days") to improve the scale's sensitivity to milder alterations, and thus increase its usefulness in subclinical populations. In study 1 (N=485), we provide evidence using Item Response Theory (IRT) that the new response option does improve the scale's psychometric quality and extends the sensitivity to the measured constructs on the lower end of the spectrum. In study 2 (N=836), we show that the refined version offers an improved sensitivity to subclinical variability in depression (indexed by the BDI-II) as compared to the original version. In conclusion, adding the "once or twice" response option is a low-cost no-downsides way of increasing the PHQ-4's sensitivity to subclinical variability, making it a tool of choice for general population research.

@@

floatsintext: true

@@

While the scale has been validated and used in the general population and non-clinical samples [@lowe20104; @hajek2020prevalence], its initial purpose was to reliably discriminate and identify potential MDD/GAD patients. This discriminative goal materializes in the scale's design and the existence of categorical cut-offs, which does not necessary entail a focus on the sensitivity to milder mood alterations. In particular, the gap between the two lowest possible answers, "Not at all" and "Several days", is quite large and possibly leaves out the possibility of more subtle occurrences. While this is not necessarily an issue in clinical and diagnostic contexts, it might lead to a sub-optimal discrimination of affective levels on the lower end of the spectrum, important for instance in the context of subclinical variability quantification.

@@

In this sample, 51 participants (6.10%) were labelled as having Depression, as indexed by the self-reported presence of MDD together with the use of a treatment (antidepressent, anxiolytic and/or therapy), and 87 participants (10.41%) were labelled as having Anxiety, as indexed by the self-reported presence of GAD or Panic Disorder, also together with the use of a treatment.

@@

The original or refined version of the PHQ-4 was followed by the BDI-II, STAI-5, IAS, and MAIA-2, presented in random order. The IAS and the MAIA-2 were included as part of another study focused on interoception, and were only used in this stud as part of data quality control checks. 
## Results

@@

The objective of this study was to test the introduction of a "Once or twice" response option to the PHQ-4 to enhance its sensitivity to milder mood fluctuations. In the first study, we showed that the new response option was used prevalently by participants and did capture a unique portion of the depression and anxiety underlying dimensions. In the second study, we showed that the refined version of the PHQ-4 was able to better differentiate lower levels of depression compared to the original version, while remaining comparable. Although the benefits of this refinement appear to be fairly minor, and particularly marked for the depression score compared to anxiety, this cost-free improvement appear useful to implement when measuring depression and anxiety using the PHQ-4 ultra-short screening questionnaire.

>>>>

  The PHQ-4 is an ultra-brief (4 items) screening questionnaire for depression and anxiety. In this brief report, we test the benefits of adding one additional response option ("Once or twice", in between "Not at all" and "Several days") to improve the scale's sensitivity to milder alterations, and thus increase its usefulness in subclinical populations. In study 1 (N=485), we provide evidence using Item Response Theory (IRT) that the new response option does improve the scale's psychometric quality and extends the sensitivity to the measured constructs on the lower end of the spectrum. In study 2 (N=836), we show that the refined version offers an improved sensitivity to subclinical variability in depression (indexed by the BDI-II) as compared to the original version. In conclusion, adding the "once or twice" response option is a low-cost way of increasing the PHQ-4's sensitivity to subclinical variability, making it a tool of choice for general population research.

@@

floatsintext: false
mask: true

@@

While the scale has been validated and used in the general population and non-clinical samples [@lowe20104; @hajek2020prevalence], its initial purpose was to reliably discriminate and identify potential MDD/GAD patients. This discriminative goal materializes in the scale's design and the existence of categorical cut-offs, which does not necessarily entail a focus on the sensitivity to milder mood alterations. In particular, the gap between the two lowest possible answers, "Not at all" and "Several days", is quite large and possibly leaves out the possibility of more subtle occurrences. While this is not necessarily an issue in clinical and diagnostic contexts, it might lead to a sub-optimal discrimination of affective levels on the lower end of the spectrum, important for instance in the context of subclinical variability quantification.

@@

In this sample, 51 participants (6.10%) were coded as having Depression, as indexed by the self-reported presence of MDD together with the use of a treatment (antidepressent, anxiolytic and/or therapy), and 87 participants (10.41%) were coded as having Anxiety, as indexed by the self-reported presence of GAD or Panic Disorder, also together with the use of a treatment.

@@

The original or refined version of the PHQ-4 was followed by the BDI-II, STAI-5, IAS, and MAIA-2, presented in random order. The IAS and the MAIA-2 were included as part of another study focused on interoception, and were only used in this stud as part of data quality control checks.

@@

The objective of this study was to test the introduction of a "Once or twice" response option to the PHQ-4 to enhance its sensitivity to milder mood fluctuations. In the first study, we showed that the new response option was used prevalently by participants and did capture a unique portion of the depression and anxiety underlying dimensions. In the second study, we showed that the refined version of the PHQ-4 was able to better differentiate lower levels of depression compared to the original version, while remaining comparable. Although the benefits of this refinement appear to be fairly minor, and particularly marked for the depression score compared to anxiety, this low-cost improvement appear useful to implement when measuring depression and anxiety using the PHQ-4 ultra-short screening questionnaire.
